---
title: "Rural Urban"
author: "Alisa Hartle"
date: "December 9, 2018"
output: html_document
---

#Overview of Urban-Rural Counties in Pennsylvania and Ohio
```{r}
source("OHmerge.R")
OH <- getOH() #FINAL OH DATA SET 
source("PAmerge.R")
PA <- getPA() #FINAL PA DATA SET 
summary(OH$ruralurban_cc)
summary(PA$ruralurban_cc)

#Ruralurban_cc measures how rural an area is; 1 indicates counties of
#1,000,000 people or more, while indicates a county with population < 2,500

#The summary statistics of these two states show that  the median PA county is more "urban" than OH in terms of population density. However,PA also has a wider variation in county urban-rural spectrum (1-9 compared with OH's 1-8). 

hist(PA$ruralurban_cc, breaks = 5)
hist(OH$ruralurban_cc, breaks =5)

#The histograms provide a little more insight. PA counties have a natural variation in population size between large, medium, and small. Ohio looks quite different. Ohio has high density populated counties like Pennyslvania, but fewer moderate-to-high denisty populated counties. This urban isolation in Ohio could be a possible predictor later on in the analysis.
```

#Distinguishing Between Counties
```{r}
pa_urban <-subset(PA, PA$ruralurban_cc == 1 | 
                 PA$ruralurban_cc == 2)
pa_suburb <- subset(PA, PA$ruralurban_cc == 3 |
                     PA$ruralurban_cc == 4 | 
                     PA$ruralurban_cc == 5)
pa_rural <- subset(PA, PA$ruralurban_cc == 6 |
                  PA$ruralurban_cc == 7 | 
                  PA$ruralurban_cc == 8|
                  PA$ruralurban_cc == 9|
                  PA$ruralurban_cc == 10)

#Out of 67 counties in Pennsylvania, 19 are rural, 21 are suburban, and 27 are urban.


oh_urban <-subset(OH, OH$ruralurban_cc == 1 | 
                 OH$ruralurban_cc == 2)
oh_suburb <- subset(OH, OH$ruralurban_cc == 3 |
                     OH$ruralurban_cc == 4 | 
                     OH$ruralurban_cc == 5)
oh_rural <- subset(OH, OH$ruralurban_cc == 6 |
                  OH$ruralurban_cc == 7 | 
                  OH$ruralurban_cc == 8|
                  OH$ruralurban_cc == 9|
                  OH$ruralurban_cc == 10)

#Out of 88 counties in Ohio, 28 are rural, 27 are suburbs, and 33 are urban.
```

#Vote Test Function
```{r}
#vote_test provides a basic framework to compare percentage of Republican voters in two different datasets 
#a and b are the datasets being compared
#c and d are the races we're examining
vote_test <- function(a, b, c, d) {
  vote_test <-t.test(a[,c]/(a[,c]+a[,d]), b[,c]/(b[,c]+b[,d]))
  return(vote_test)
}

```


#Hypothesis 2a: rural counties are more likely to vote Republican and urban counties are more likely to vote Democratic
```{r}
vote_test(a=pa_rural,b=pa_urban, c="Dcongress18",d="Rcongress18")
vote_test(a=pa_rural,b=pa_urban, c="demhouse16",d="rephouse16")
vote_test(a=oh_rural,b=oh_urban, c="Dcongress18",d="Rcongress18")
vote_test(a=oh_rural,b=oh_urban, c="demhouse16",d="rephouse16")
```
#Hypothesis 2b: Which rural or urban counties have the lowest margins between Republican and Democrat? Are there any predictors for these counties?
```{r}
#diff calculates the difference in % of the vote between Republican and Democratic candidate
diff <- function(df, rep, dem) {
  diff <- abs((df[,rep]/(df[,rep]+df[,dem]))-(df[,dem]/(df[,rep]+df[,dem])))
}

summary(diff(df=pa_rural, "Rcongress18","Dcongress18"))
summary(diff(df=pa_suburb, "Rcongress18","Dcongress18"))
summary(diff(df=pa_urban, "Rcongress18","Dcongress18"))

PA_cong_marg <-subset(PA,diff(df=PA, "Rcongress18","Dcongress18")<.05)
OH_cong_marg <-subset(OH,diff(df=OH, "Rcongress18","Dcongress18")<.05)

#The number of counties with margins of less than 5% are so few that any analysis of these subsets would be insignificant. We're beginning to see the limits of this dataset.
```

#Hypothesis 2c: Rural and urban counties are most likely to vote “straight ticket” by electing the same party in multiple offices.
```{r}

```

#Hypothesis 2d: For rural/urban districts that do not vote straight ticket, we’re hypothesising this has been a consistent trend.
```{r}

```

#Hypothesis 2e: Suburban districts are most likely to “flip” between election cycles. Our hypothesis is that counties that have shifted politically in 2018 have shown similar behavior in the past.
```{r}

```

#Hypothesis 2f: Additional demographic indicators such as education, race, or income could predict which suburban districts are most likely to flip in midterm elections.
```{r}

```

